Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-27 submit-2/paper.tex
added resubmission folder

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small structure-activity
  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise combinations of molecular
  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. While the models do not exhibit
  very high levels of accuracy, our results indicate that they are able to prioritize molecules in
  terms of their ability to activity cliffs, thus serving as a tool to prospectively identify
  activity cliffs..
\end{abstract}

@@

\bibliography{paper}

>>>>

\maketitle
\begin{abstract}
  The notion of activity cliffs is an intuitive approach to characterizing structural features that
  play a key role in modulating biological activity of a molecule. A variety of methods have been
  described to quantitatively characterize activity cliffs, such as SALI and SARI. However, these
  methods are primarily retrospective in nature; highlighting cliffs that are already present in the
  dataset. The current study focuses on employing a pairwise characterization of a dataset to train
  a model to predict whether a new molecule will exhibit an activity cliff with one or more members
  of the dataset. The approach is based on predicting a value for pairs of objects rather than the
  individual objects themselves (and thus allows for robust models even for small structure-activity
  relationship datasets). We extracted structure-activity data for several ChEMBL assays and
  developed random forest models to predict SALI values, from pairwise combinations of molecular
  descriptors. The models exhibited reasonable RMSE's though, surprisingly, performance on the more
  significant cliffs tended to be better than on the lesser ones. While the models do not exhibit
  very high levels of accuracy, our results indicate that they are able to prioritize molecules in
  terms of their ability to activity cliffs, thus serving as a tool to prospectively identify
  activity cliffs.
\end{abstract}

@@

%%\bibliography{paper}

\providecommand{\refin}[1]{\\ \textbf{Referenced in:} #1}
\begin{thebibliography}{10}

\bibitem{Johnson:1990ys}
Johnson,~M.;\ \ Maggiora,~G. \textit{Concepts and Applications of Molecular
  Similarity;} John Wiley \& Sons: New York, 1990.

\bibitem{Maggiora:2006aa}
Maggiora,~G.~M.  On Outliers and Activity Cliffs--Why {QSAR} Often
  Disappoints..  \textit{J.~Chem.~Inf.~Model.} \textbf{2006,} \textsl{46,}
  1535--1535.

\bibitem{Leach:2006aa}
Leach,~A.;\ \ Jones,~H.;\ \ Cosgrove,~D.;\ \ Kenny,~P.;\ \ Ruston,~L.;\ \
  MacFaul,~P.;\ \ Wood,~J.;\ \ Colclough,~N.;\ \ Law,~B.  Matched Molecular
  Pairs as a Guide in the Optimization of Pharmaceutical Properties; a Study of
  Aqueous Solubility, Plasma Protein Binding and Oral Exposure.
  \textit{J.~Med.~Chem.} \textbf{2006,} \textsl{49,} 6672--6682.

\bibitem{Maggiora:2011kx}
Shanmugasundaram,~V.;\ \ Maggiora,~G.  Characterizing Property and Activity
  Landscapes Using an Information-Theoretic Approach.   In  \textit{CINF-032.
  222nd ACS National Meeting, Chicago, IL, United States}; American Chemical
  Society: Washington, D.C., 2001.

\bibitem{Guha:2008aa}
Guha,~R.;\ \ Van~Drie,~J.  The Structure-Activity Landscape Index:
  {I}dentifying and Quantifying Activity-Cliffs.  \textit{J.~Chem.~Inf.~Model.}
  \textbf{2008,} \textsl{48,} 646--658.

\bibitem{Peltason:2007aa}
Peltason,~L.;\ \ Bajorath,~J.  SAR Index: Quantifying the Nature of
  Structure-Activity Relationships.  \textit{J.~Med.~Chem.} \textbf{2007,}
  \textsl{50,} 5571--5578.

\bibitem{Wawer:2008aa}
Wawer,~M.;\ \ Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.
  Structure-Activity Relationship Anatomy by Network-like Similarity Graphs and
  Local Structure-Activity Relationship Indices.  \textit{J.~Med.~Chem.}
  \textbf{2008,} \textsl{51,} 6075--6084.

\bibitem{Vogt:2011bs}
Vogt,~M.;\ \ Huang,~Y.;\ \ Bajorath,~J.  From Activity Cliffs to Activity
  Ridges: {I}nformative Data Structures for SAR Analysis.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,} 1848--1856.

\bibitem{Dimova:2011fk}
Dimova,~D.;\ \ Wawer,~M.;\ \ Wassermann,~A.~M.;\ \ Bajorath,~J.  Design of
  Multitarget Activity Landscapes That Capture Hierarchical Activity Cliff
  Distributions.  \textit{J.~Chem.~Inf.~Model.} \textbf{2011,} \textsl{51,}
  258--266.

\bibitem{Wawer:2010fv}
Wawer,~M.;\ \ Bajorath,~J.  Similarity-potency trees: {A} method to search for
  {SAR} information in compound data sets and derive {SAR} rules.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1395--1409.

\bibitem{Hu:2010kl}
Hu,~Y.;\ \ Bajorath,~J.  Molecular Scaffolds with High Propensity to Form
  Multi-Target Activity Cliffs.  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,}
  \textsl{50,} 500--510.

\bibitem{Wassermann:2010oq}
Wassermann,~A.~M.;\ \ Bajorath,~J.  Chemical Substitutions that Introduce
  Activity Cliffs Across Different Compound Classes and Biological Targets.
  \textit{J.~Chem.~Inf.~Model.} \textbf{2010,} \textsl{50,} 1248-1256.

\bibitem{Peltason:2009vn}
Peltason,~L.;\ \ Weskamp,~N.;\ \ Teckentrup,~A.;\ \ Bajorath,~J.  Exploration
  of Structure-Activity Relationship Determinants in Analogue Series.
  \textit{J.~Med.~Chem.} \textbf{2009,} \textsl{52,} 3212--3224.

\bibitem{Scheiber:2009fk}
Scheiber,~J.;\ \ Jenkins,~J.~L.;\ \ Bender,~A.;\ \ Milik,~M.;\ \ Mikhalov,~D.;\
  \ Sukuru,~S.;\ \ Cornett,~B.;\ \ Whitebread,~S.;\ \ Laszlo,~U.;\ \
  Davies,~J.;\ \ Glick,~M.  {SPREAD} - exploiting chemical features that cause
  differential activity behavior.  \textit{Stat.~Anal.~Data Mining}
  \textbf{2009,} \textsl{2,} 115-122.

\bibitem{Cavalli:2002aa}
Cavalli,~A.;\ \ Poluzzi,~E.;\ \ De~Ponti,~F.;\ \ Recanatini,~M.  Toward a
  Pharmacophore for Drugs Inducing the Long {QT} Syndrome: Insights from a
  {CoMFA} Study of {HERG K+} Channel Blockers.  \textit{J.~Med.~Chem.}
  \textbf{2002,} \textsl{45,} 3844--3853.

\bibitem{Costanzo:2005uq}
Costanzo,~M.~J.;\ \ Almond,~Jr,~H.~R.;\ \ Hecker,~L.~R.;\ \ Schott,~M.~R.;\ \
  Yabut,~S.~C.;\ \ Zhang,~H.-C.;\ \ Andrade-Gordon,~P.;\ \ Corcoran,~T.~W.;\ \
  Giardino,~E.~C.;\ \ Kauffman,~J.~A.;\ \ Lewis,~J.~M.;\ \ de~Garavilla,~L.;\ \
  Haertlein,~B.~J.;\ \ Maryanoff,~B.~E.  In-depth study of tripeptide-based
  alpha-ketoheterocycles as inhibitors of thrombin. {E}ffective utilization of
  the {S1'} subsite and its implications to structure-based drug design.
  \textit{J.~Med.~Chem.} \textbf{2005,} \textsl{48,} 1984--2008.

\bibitem{Kalla:2006kx}
Kalla,~R.~V.;\ \ Elzein,~E.;\ \ Perry,~T.;\ \ Li,~X.;\ \ Palle,~V.;\ \
  Varkhedkar,~V.;\ \ Gimbel,~A.;\ \ Maa,~T.;\ \ Zeng,~D.;\ \ Zablocki,~J.
  Novel 1,3-disubstituted 8-(1-benzyl-1H-pyrazol-4-yl) xanthines: high affinity
  and selective A2B adenosine receptor antagonists.  \textit{J.~Med.~Chem.}
  \textbf{2006,} \textsl{49,} 3682--3692.

\bibitem{Dai:2007kl}
Dai,~Y. \textit{et al.}\   Discovery of
  N-(4-(3-amino-1H-indazol-4-yl)phenyl)-N'-(2-fluoro-5-methylphenyl)urea
  (ABT-869), a 3-aminoindazole-based orally active multitargeted receptor
  tyrosine kinase inhibitor.  \textit{J.~Med.~Chem.} \textbf{2007,}
  \textsl{50,} 1584--1597.

\bibitem{Steinbeck:2006aa}
Steinbeck,~C.;\ \ Hoppe,~C.;\ \ Kuhn,~S.;\ \ Floris,~M.;\ \ Guha,~R.;\ \
  Willighagen,~E.  Recent Developments of the Chemistry Development Kit ({CDK})
  - An Open-Source {Java} Library for Chemo- and Bioinformatics.
  \textit{Curr.~Pharm.~Des.} \textbf{2006,} \textsl{12,} 2110--2120.

\bibitem{Steinbeck:2003bh}
Steinbeck,~C.;\ \ Han,~Y.~Q.;\ \ Kuhn,~S.;\ \ Horlacher,~O.;\ \ Luttmann,~E.;\
  \ Willighagen,~E.  The Chemistry Development Kit ({CDK}): An Open-Source
  {J}ava Library for Chemo- and Bioinformatics.
  \textit{J.~Chem.~Inf.~Comput.~Sci.} \textbf{2003,} \textsl{43,} 493--500.

\bibitem{Breiman:1984aa}
Breiman,~L.;\ \ Friedman,~J.;\ \ Olshen,~R.;\ \ Stone,~C.
  \textit{Classification and Regression Trees;} Chapman \& Hall/CRC: Boca
  Raton, FL, 1984.

\bibitem{Breiman:2001nx}
Breiman,~L.  Statistical Modeling: Two Cultures.  \textit{Stat.~Sci.}
  \textbf{2001,} \textsl{16,} 199--231.

\bibitem{r}
{R Development Core Team}, ``{R}: {A} Language and Environment for Statistical
  Computing'',  R Foundation for Statistical Computing, Vienna, Austria, 2008
  {ISBN} 3-900051-07-0.

\bibitem{Rucker:2007aa}
Rucker,~C.;\ \ Rucker,~G.;\ \ Meringer,~M.  y-Randomization and Its Variants in
  {QSPR/QSAR}.  \textit{J.~Chem.~Inf.~Model.} \textbf{2007,} \textsl{47,}
  2345--2357.

\end{thebibliography}
